The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-11-27 script/en/chapters/primes.tex
Script[de+en]: Many small changes, e.g. Sage samples collected in files, use of iload; described how to deal with Sage notebook using interact; substituted href by url.

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signature!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

The following table contains the current record primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \href{http://primes.utm.edu/largest.html}
  {\texttt{http://primes.utm.edu/largest.html}}
and at
     \href{http://primes.utm.edu/mersenne/index.html}
  {\texttt{http://primes.utm.edu/mersenne/index.html}}.
}:

@@

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of July 2009)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\footnotetext{\index{Number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\href{http://www.seventeenorbust.com}{\texttt{http://www.seventeenorbust.com}})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its
task sometime.

@@

The famous Polish mathematician Waclaw Sierpinski (1882 to 1969) proved in
1960, that there exist infinitely many odd integers k, which fulfill the
following property: For all Sierpinski numbers k it is true: All numbers $N = k \cdot 2^{n}+1$ are composite for all integers $n>=1$ (Sierpinski's Composite Number Theorem, \href{http://mathworld.wolfram.com/SierpinskisCompositeNumberTheorem.html}{\texttt{http://mathworld.wolfram.com/SierpinskisCompositeNumberTheorem.html}}).

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \href{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}
   {\texttt{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}}. \\
Also see:
      \href{http://www.utm.edu/}
   {\texttt{http://www.utm.edu/}}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 43,112,609; ~ 42,643,801.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$47$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +25 pt
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \href{http://www.mersenne.org/status.htm}
   {\texttt{http://www.mersenne.org/status.htm}}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}. }.

@@

Right now (July 2009) all prime exponents smaller than $ 18,000,989 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \href{http://www.mersenne.org/report_milestones}
   {\texttt{http://www.mersenne.org/report\_milestones/}}.
                                  }:
So we can
be certain, that this is really the 39th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime).

@@

The GIMPS project (Great Internet Mersenne Prime Search)\index{GIMPS} was
founded in 1996 by George Woltman\index{Woltman, George} to search for new
largest Mersenne primes 
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).
Further explanations about this number type can be found under
\hyperlink{MersenneNumbers02}{Mersenne numbers} and 
\hyperlink{MersenneNumbers01}{Mersenne primes}.

@@

The following table contains these Mersenne record primes\footnote{%
An up-to-date version can be found in the internet at
     \href{http://www.mersenne.org/history.htm}
  {\texttt{http://www.mersenne.org/history.htm}}.
}$^,$\footnote{%
Always, when a new record is published in the respective forums the same
and often ironic discussions start: Does this kind of research have a deeper
sense? Can this result be applied for anything useful?
The answer is, that we don't know it yet. In fundamental research one
cannot see at once whether and how it brings mankind forward.
%But this is the normal case with basic research, that one cannot see at once, 
}:

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsection{Challenge of the Electronic Frontier Foundation (EFF)}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organization EFF (Electronic Frontier Foundation) using the means of an 
unknown donor. The participants are rewarded with a total of 500,000 USD if
they find the longest prime number. In promoting this project, the unknown
donor is not looking for the quickest computer, but rather wants to draw 
people's attention to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/awards/coop}{\tt http://www.eff.org/awards/coop}}

@@

% be_2005_UPDATEN_if-new-mersenne-prime-appears
For the next prize of 100,000 USD offered by EFF for a proven prime with more
than 10 million decimal digits, Edson Smith qualified, who found the number
$ 2^{43,112,609}-1 $ within the GIMPS project.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

Detailed descriptions of tests for finding out whether a number is prime
can be found on Web sites such as:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/mersenne.shtml}
               {\texttt{http://www.utm.edu/research/primes/mersenne.shtml}} \\
          \href{http://www.utm.edu/research/primes/prove/index.html}
               {\texttt{http://www.utm.edu/research/primes/prove/index.html}} 
\end{itemize}

@@

Ideas for this can be found at
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}
          {\tt http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.
\end{itemize}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsection{Mersenne numbers \texorpdfstring{$f(n) = 2^n - 1$  \quad for $ n $ prime}
	   			      {f(n) = 2\^{}n - 1 for n prime}}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits
    are known (\hyperlink{M-39}{M-39}\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}      % be_2005_UPDATEN_if-new-mersenne-prime-appears

@@

   This first generalization of the Mersenne
   numbers\index{Mersenne!number!generalized}
   creates the so called Proth numbers\index{Number!Proth}.
   There are (for small $k$) extremely quick prime number tests
   (see \cite{pr:Knuth1981}). This can be performed in practice using software
   such as the Proths software from Yves Gallot\index{Gallot, Yves}:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
            {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

   Details of this can be found at:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.cerias.purdue.edu/homes/ssw/cun}
            {\tt http://www.cerias.purdue.edu/homes/ssw/cun}}
\end{itemize}

@@

    Within the project ``Distributed Search for Fermat Number Dividers''
    offered by Leonid Durman there is also progress in finding new monster
    primes: 
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.fermatsearch.org/}
       {\tt http://www.fermatsearch.org/}}\\
       This website links to other web pages in Russian, Italian and German.
\end{itemize}
    The discovered factors can be compound integers or primes.

@@

    More details are in
    \vspace{-10pt}
    \begin{itemize}
      \item[] \href{http://www.fermatsearch.org/history/cosgrave_record.htm/}
          {\texttt{http://www.fermatsearch.org/history/cosgrave\_record.htm/}}
    \end{itemize}

%    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
%           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).
% M.E. ist die Zahl auf seiner Webseite zu groﬂ, da M-39 "nur" 4 Mio. Stellen hat !
%    Heute kennt man alle Fermatschen Primzahlen mit bis zu 2.000.000.000 ??? 
%    Dezimalstellen. \\

@@

    More details are in
    \vspace{-10pt}
    \begin{itemize}
      \item[] \href{http://primes.utm.edu/top20/page.php?id=12}
              {\texttt{http://primes.utm.edu/top20/page.php?id=12}} 
    \end{itemize}

@@

Under the heading {\em How Many Primes Are There} at
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/howmany.shtml}
               {\tt http://www.utm.edu/research/primes/howmany.shtml}
\end{itemize}
\vspace{-10pt}
you will find numerous other details.

Using the following Web site:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}
               {\tt http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}
\end{itemize}
\vspace{-10pt}
you can easily determine $PI(x)$.

The \textbf{distribution} of primes displays several irregularities for which
no ``system'' has yet been found: On the one hand, many occur closely together,
like $2$ and $3,$ $ 11$ and $13, $ $ 809$ and $811$, on the other hand large
gaps containing no primes also occur. For example, no primes lie between $113$
and $127, $ $ 293$ and $307, $ $317$ and $331, $ $ 523$ and $541, $ $ 773$
and $787, $ $ 839$ and $853$ as well as between $887$ and $907$. \\
For details, please see: 
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/notes/gaps.html}
               {\tt http://www.utm.edu/research/primes/notes/gaps.html}
\end{itemize}

@@

  \vskip +6 pt      
  \item \hypertarget{link-Arithmetic-sequence-of-primes}{}
     There are arithmetic prime sequences\index{Prime sequence!arithmetic}
     of arbitrary length\footnote{%
     Sources: \\
     - \href{http://primes.utm.edu/glossary/page.php?sort=ArithmeticSequence}
     {\texttt{http://primes.utm.edu/glossary/page.php?sort=ArithmeticSequence}}
       Original source\\
     - German magazine GEO 10 / 2004: ``Experiment mit Folgen''\\    %``xx''
     - \href{http://www.faz.net}
       {\texttt{http://www.faz.net}} ``Hardys Vermutung -- Primzahlen ohne 
                                       Ende'' by Heinrich Hemme (July 06, 2004)
     }.

@@

     {\bf Further restriction:} If you look at arithmetic prime sequences, 
     which fulfill the {\em additional} requirement, that all primes are
     consecutive, then its getting even more complicated. At the website
     of Chris Caldwell\footnote{%
     \href{http://primes.utm.edu/glossary/page.php?sort=ArithmeticSequence}
     {\texttt{http://primes.utm.edu/glossary/page.php?sort=ArithmeticSequence}}
     } 
     you can find further hints: the longest known arithmetic prime sequence, 
     consisting only of directly consecutive primes, has a length of $10$ and
     the distance is $210 = 2 * 3 * 5 * 7$. \\

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Unproven statements / conjectures about primes}
\begin{itemize}
  \item Christian Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.
     Computers have verified\footnote{%
     It is generally accepted today, that the \textbf{Goldbach Conjecture} is true, 
     i.~e.\ valid for all even natural numbers greater than $2$.
     In 1999, mathematician J\"org Richstein\index{Richstein 1999}
     from the computer sciences institute at the University of Giessen,
     studied even numbers up to 400 billion ($4*10^{14}$) and found no
     contradictory example (\cite{pr:Richstein1999}).\\
     In the meantime further progress was made: see\\
     \url{http://www.ieeta.pt/~tos/goldbach.html},\\
     \url{http://en.wikipedia.org/wiki/Goldbach's\_conjecture},\\
     \url{http://primes.utm.edu/glossary/page.php/GoldbachConjecture.html}.\\      
     Nevertheless, this does not provide us with general proof.\\
     The fact is that despite all efforts, Goldbach's conjecture has to 
     date not been proven. This leads one to believe that since the pioneer
     work of the Austrian mathematician Kurt G\"odel \index{G\""odel, Kurt}
     is well-known, not every true mathematical theorem is provable (see
     \href{http://www.mathematik.ch/mathematiker/goedel.html}
     {http://www.mathematik.ch/mathematiker/goedel.html}).
     Perhaps Goldbach's conjecture was correct, but in any case the proof
     will never be found. Conversely, that will presumably also remain 
     unproven.}
     the Goldbach conjecture for all even numbers up to
     $15*10^{17}$ (as at July 2009) but no general proof has yet been
     found\footnote{%
     The English publisher {\em Faber} and the American publisher 
     {\em Bloomsbury} issued in 2000 the 1992 published book 
     ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
     It's the story of an old maths professor who fails to prove a more 
     than 250 year old puzzle. To boost the sales figures the English and
     American publishers have offered a prize of 1 million USD, if someone
     can prove the conjecture -- which should be published by 2004 in a 
     well-known mathematical journal.\\
     Surprisingly only British and American citizens are allowed to 
     participate.\\
     The theorem which has come closest so far to Goldbach's conjecture
     was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
     to understand:
     Each even integer greater than $2$ is the sum of one prime and of 
     the product of two primes. E.g.: $20=5+3*5.$\\
     Most of the research about the Goldbach conjecture is collected in 
     the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World 
     scientific Series in Pure Maths, Vol. 4.\\
     Especially this conjecture makes it clear, that even today we do not
     have a complete understanding of the deeper connections between 
     addition and multiplication of natural numbers. }.

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsection[Quaint and interesting things around primes]
              {Quaint and interesting things around primes\footnotemark}
    \footnotetext{%
Further curious things about primes may be found at:\\
- \href{http://primes.utm.edu/curios/home.php}
   {\tt http://primes.utm.edu/curios/home.php}\\
- \href{http://www.primzahlen.de/files/theorie/index.htm}
   {\tt http://www.primzahlen.de/files/theorie/index.htm}.
}
\label{HT-Quaint-curious-Primes-usage}

>>>>

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signature!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\
(see also: \url{http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

The following table contains the current record primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \url{http://primes.utm.edu/largest.html}
and at
     \url{http://primes.utm.edu/mersenne/index.html}.
}:

@@

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of July 2009)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\footnotetext{\index{Number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its
task sometime.

@@

The famous Polish mathematician Waclaw Sierpinski (1882 to 1969) proved in
1960, that there exist infinitely many odd integers k, which fulfill the
following property: For all Sierpinski numbers k it is true: All numbers $N = k \cdot 2^{n}+1$ are composite for all integers $n>=1$ (Sierpinski's Composite Number Theorem, \url{http://mathworld.wolfram.com/SierpinskisCompositeNumberTheorem.html}).

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html} \\
Also see:
      \url{http://www.utm.edu/}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 43,112,609; ~ 42,643,801.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$47$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +25 pt
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/status.htm}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}. }.

@@

Right now (July 2009) all prime exponents smaller than $ 18,000,989 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \url{http://www.mersenne.org/report_milestones}.
                                  }:
So we can
be certain, that this is really the 39th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime).

@@

The GIMPS project (Great Internet Mersenne Prime Search)\index{GIMPS} was
founded in 1996 by George Woltman\index{Woltman, George} to search for new
largest Mersenne primes ({\url{http://www.mersenne.org}}).
Further explanations about this number type can be found under
\hyperlink{MersenneNumbers02}{Mersenne numbers} and 
\hyperlink{MersenneNumbers01}{Mersenne primes}.

@@

The following table contains these Mersenne record primes\footnote{%
An up-to-date version can be found in the internet at
     \url{http://www.mersenne.org/history.htm}.
}$^,$\footnote{%
Always, when a new record is published in the respective forums the same
and often ironic discussions start: Does this kind of research have a deeper
sense? Can this result be applied for anything useful?
The answer is, that we don't know it yet. In fundamental research one
cannot see at once whether and how it brings mankind forward.
%But this is the normal case with basic research, that one cannot see at once, 
}:

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsection{Challenge of the Electronic Frontier Foundation (EFF)}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organization EFF (Electronic Frontier Foundation) using the means of an 
unknown donor. The participants are rewarded with a total of 500,000 USD if
they find the longest prime number. In promoting this project, the unknown
donor is not looking for the quickest computer, but rather wants to draw 
people's attention to the opportunities offered by {\em cooperative networking} \\
{\url{http://www.eff.org/awards/coop}}

@@

% be_2005_UPDATEN_if-new-mersenne-prime-appears
For the next prize of 100,000 USD offered by EFF for a proven prime with more
than 10 million decimal digits, Edson Smith qualified, who found the number
$ 2^{43,112,609}-1 $ within the GIMPS project.
%{\url{http://www.octocad.demon.co.uk/mersenne/prime.htm }}.

@@

Detailed descriptions of tests for finding out whether a number is prime
can be found on Web sites such as:
\vspace{-10pt}
\begin{itemize}
  \item[] \url{http://www.utm.edu/research/primes/mersenne.shtml} \\
          \url{http://www.utm.edu/research/primes/prove/index.html}
\end{itemize}

@@

Ideas for this can be found at
\vspace{-10pt}
\begin{itemize}
  \item[] {\url{http://www.utm.edu/research/primes/notes/faq/p_n.html}}.
\end{itemize}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsection{Mersenne numbers \texorpdfstring{$f(n) = 2^n - 1$  \quad for $ n $ prime}
	   			      {f(n) = 2\^{}n - 1 for n prime}}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits
    are known (\hyperlink{M-39}{M-39}\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\url{http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}      % be_2005_UPDATEN_if-new-mersenne-prime-appears

@@

   This first generalization of the Mersenne
   numbers\index{Mersenne!number!generalized}
   creates the so called Proth numbers\index{Number!Proth}.
   There are (for small $k$) extremely quick prime number tests
   (see \cite{pr:Knuth1981}). This can be performed in practice using software
   such as the Proths software from Yves Gallot\index{Gallot, Yves}:
\vspace{-10pt}
\begin{itemize}
  \item[] {\url{http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

   Details of this can be found at:
\vspace{-10pt}
\begin{itemize}
  \item[] {\url{http://www.cerias.purdue.edu/homes/ssw/cun}}
\end{itemize}

@@

    Within the project ``Distributed Search for Fermat Number Dividers''
    offered by Leonid Durman there is also progress in finding new monster
    primes: 
\vspace{-10pt}
\begin{itemize}
  \item[] {\url{http://www.fermatsearch.org/}}\\
       This website links to other web pages in Russian, Italian and German.
\end{itemize}
    The discovered factors can be compound integers or primes.

@@

    More details are in
    \vspace{-10pt}
    \begin{itemize}
      \item[] \url{http://www.fermatsearch.org/history/cosgrave_record.htm/}
    \end{itemize}

%    ({\url{http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).
% M.E. ist die Zahl auf seiner Webseite zu groﬂ, da M-39 "nur" 4 Mio. Stellen hat !
%    Heute kennt man alle Fermatschen Primzahlen mit bis zu 2.000.000.000 ??? 
%    Dezimalstellen. \\

@@

    More details are in
    \vspace{-10pt}
    \begin{itemize}
      \item[] \url{http://primes.utm.edu/top20/page.php?id=12}
    \end{itemize}

@@

Under the heading {\em How Many Primes Are There} at
\vspace{-10pt}
\begin{itemize}
  \item[] \url{http://www.utm.edu/research/primes/howmany.shtml}
\end{itemize}
\vspace{-10pt}
you will find numerous other details.

Using the following Web site:
\vspace{-10pt}
\begin{itemize}
  \item[] \url{http://www.math.Princeton.EDU/~arbooker/nthprime.html}
\end{itemize}
\vspace{-10pt}
you can easily determine $PI(x)$.

The \textbf{distribution} of primes displays several irregularities for which
no ``system'' has yet been found: On the one hand, many occur closely together,
like $2$ and $3,$ $ 11$ and $13, $ $ 809$ and $811$, on the other hand large
gaps containing no primes also occur. For example, no primes lie between $113$
and $127, $ $ 293$ and $307, $ $317$ and $331, $ $ 523$ and $541, $ $ 773$
and $787, $ $ 839$ and $853$ as well as between $887$ and $907$. \\
For details, please see: 
\vspace{-10pt}
\begin{itemize}
  \item[] \url{http://www.utm.edu/research/primes/notes/gaps.html}
\end{itemize}

@@

  \vskip +6 pt      
  \item \hypertarget{link-Arithmetic-sequence-of-primes}{}
     There are arithmetic prime sequences\index{Prime sequence!arithmetic}
     of arbitrary length\footnote{%
     Sources: \\
     - \url{http://primes.utm.edu/glossary/page.php?sort=ArithmeticSequence}
       Original source\\
     - German magazine GEO 10 / 2004: ``Experiment mit Folgen''\\    %``xx''
     - \url{http://www.faz.net}
       ``Hardys Vermutung -- Primzahlen ohne Ende'' by Heinrich Hemme (July 06, 2004)
     }.

@@

     {\bf Further restriction:} If you look at arithmetic prime sequences, 
     which fulfill the {\em additional} requirement, that all primes are
     consecutive, then its getting even more complicated. At the website
     of Chris Caldwell\footnote{%
     \url{http://primes.utm.edu/glossary/page.php?sort=ArithmeticSequence}
     } 
     you can find further hints: the longest known arithmetic prime sequence, 
     consisting only of directly consecutive primes, has a length of $10$ and
     the distance is $210 = 2 * 3 * 5 * 7$. \\

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Unproven statements / conjectures about primes}
\begin{itemize}
  \item Christian Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.
     Computers have verified\footnote{%
     It is generally accepted today, that the \textbf{Goldbach Conjecture} is true, 
     i.~e.\ valid for all even natural numbers greater than $2$.
     In 1999, mathematician J\"org Richstein\index{Richstein 1999}
     from the computer sciences institute at the University of Giessen,
     studied even numbers up to 400 billion ($4*10^{14}$) and found no
     contradictory example (\cite{pr:Richstein1999}).\\
     In the meantime further progress was made: see\\
     \url{http://www.ieeta.pt/~tos/goldbach.html},\\
     \url{http://en.wikipedia.org/wiki/Goldbach's\_conjecture},\\
     \url{http://primes.utm.edu/glossary/page.php/GoldbachConjecture.html}.\\      
     Nevertheless, this does not provide us with general proof.\\
     The fact is that despite all efforts, Goldbach's conjecture has to 
     date not been proven. This leads one to believe that since the pioneer
     work of the Austrian mathematician Kurt G\"odel \index{G\""odel, Kurt}
     is well-known, not every true mathematical theorem is provable (see
     \url{http://www.mathematik.ch/mathematiker/goedel.html}).
     Perhaps Goldbach's conjecture was correct, but in any case the proof
     will never be found. Conversely, that will presumably also remain 
     unproven.}
     the Goldbach conjecture for all even numbers up to
     $15*10^{17}$ (as at July 2009) but no general proof has yet been
     found\footnote{%
     The English publisher {\em Faber} and the American publisher 
     {\em Bloomsbury} issued in 2000 the 1992 published book 
     ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
     It's the story of an old maths professor who fails to prove a more 
     than 250 year old puzzle. To boost the sales figures the English and
     American publishers have offered a prize of 1 million USD, if someone
     can prove the conjecture -- which should be published by 2004 in a 
     well-known mathematical journal.\\
     Surprisingly only British and American citizens are allowed to 
     participate.\\
     The theorem which has come closest so far to Goldbach's conjecture
     was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
     to understand:
     Each even integer greater than $2$ is the sum of one prime and of 
     the product of two primes. E.g.: $20=5+3*5.$\\
     Most of the research about the Goldbach conjecture is collected in 
     the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World 
     scientific Series in Pure Maths, Vol. 4.\\
     Especially this conjecture makes it clear, that even today we do not
     have a complete understanding of the deeper connections between 
     addition and multiplication of natural numbers. }.

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsection[Quaint and interesting things around primes]
              {Quaint and interesting things around primes\footnotemark}
    \footnotetext{%
Further curious things about primes may be found at:\\
- \url{http://primes.utm.edu/curios/home.php}\\
- \url{http://www.primzahlen.de/files/theorie/index.htm}.
}
\label{HT-Quaint-curious-Primes-usage}
